---
title: "Statistics 533: Final Project - Survival Analysis for Malignant Melanoma"
author: "Yitao Guan,Seoungyeon  Shim, Duc Truong, Meraj Khandakar,Jingxuan Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Structured and Components

```{r, include=FALSE}
library(survival)
library(forcats)
library(ranger)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(finalfit)
library(cmprsk)
library(MASS)

path_file <- "Melanoma.csv"
melanoma<-read.csv(path_file,header=TRUE)
```
#### time 
Survival time in days since the operation, possibly censored.
  
#### status 
  The patients status at the end of the study. 1 indicates that they had died 
  from melanoma, 2 indicates that they were still alive and 3 indicates that 
  they had died from causes unrelated to their melanoma.
  
#### sex 
  The patients sex; 1=male, 0=female.
  
#### age 
  Age in years at the time of the operation.
  
#### year 
  Year of operation.
  
#### thickness 
  Tumour thickness in mm.
  
#### ulcer 
  Indicator of ulceration; 1=present, 0=absent."
  
Sumary of statistics for our components.

```{r, echo=FALSE}
ff_glimpse(melanoma)
```

Constructing our data frame.
We define 3 groups of classification for failure time:
- General mortality where 2 is alive, and  1/3 are death.
- MM mortality comparing with 1 die by MM, 2/3 are alive.
- Censoring adjust for competing risks 2 is alive, 1 is death by MM adjusting for 3.
```{r}
melanoma.df = melanoma %>%mutate(
  
    # general mortality where 2 is alive, and  1/3 are death
    status_general = ifelse(melanoma$status == 2, 0, 1), 
    
    # MM mortality comparing with 1 die by MM, 2/3 are alive
    status_mm = case_when(melanoma$status==3 ~  0, melanoma$status==2 ~ 0,melanoma$status==1 ~ 1),
    
    # Censoring adjust for competing risks 2 is alive, 1 is death by MM adjusting for 3
    status_cen = case_when(melanoma$status == 2 ~ 0, melanoma$status == 1 ~ 1, TRUE ~ 2),
    
    # labeling components for plots and tables
    thickness = ff_label(melanoma$thickness, "Tumour thickness (mm)"),
    age = ff_label(melanoma$age, "Age (Days)"), 
    sex = factor(melanoma$sex) %>% fct_recode("Male" = "1", "Female" = "0") %>% ff_label("Gender"),
    ulcer = factor(melanoma$ulcer) %>% fct_recode("No" = "0","Yes" = "1") %>% ff_label("Ulceration")
)
```
## Kaplan Meier Analysis
```{r}
km_fit <- survfit(
  Surv(time, status_general)~1, 
  data=melanoma.df
)
plot(km_fit, xlab="Days", ylab="Cummulative Probability of survival over time for MM" ,main = "Kaplan Meier plot for the survival of Maglinant Melanoma(MM)")

```

<br/>

## KM fit adjusted for gender

```{r}
km_fit_gender <- survfit(
  Surv(time, status_general)~sex, 
  data=melanoma.df
)

xlabel = "Days"
ylabel = "Cummulative Probability of survival over time for MM"
plot_title = "Kaplan Meier plot for the survival of Maglinant Melanoma(MM) by gender"
km_gender_plot <- autoplot(km_fit_gender)
km_gender_plot <- km_gender_plot + 
  ggtitle(plot_title) +
  labs(x = xlabel, y = ylabel) +
  guides(fill=FALSE) +
  labs(colour = "Gender") +
  scale_color_manual(labels = c("Male", "Female"), values = c(1, 2))
print(km_gender_plot)
```
<br/>

## KM fit adjusted for Ulceration

```{r}
km_fit_ulcer <- survfit(
  Surv(time, status_general)~ulcer, 
  data=melanoma.df
)

plot_title = "Kaplan Meier plot for the survival of Maglinant Melanoma(MM) by ulceration"
km_ulcer_plot <- autoplot(km_fit_ulcer)
km_ulcer_plot <- km_ulcer_plot + 
  ggtitle(plot_title) +
  labs(x = xlabel, y = ylabel) +
  guides(fill=FALSE) +
  labs(colour = "Ulceration") +
  scale_color_manual(labels = c("Ulcer Present", "Ulcer Absent"), values = c(1, 2))
print(km_ulcer_plot)
```

## KM fit adjusted for ages 40,50,60,70 

"In the U.S., the lifetime risk of developing melanoma was estimated to be about 1 in 28 for Caucasians in the year 2020. The incidence of melanoma is relatively low in dark-skinned populations. Overall, the lifetime risk of developing melanoma is greater for men than women; however, for those under age 50, the inverse is true. The incidence rate begins to rise after puberty, increases until the age of 65 to 70 years, and then decreases with the average age at diagnosis being around 65."- [Munich Re Life US] <a href="https://www.munichre.com/content/dam/munichre/marc/pdf/malignant-melanoma-may-2020.pdf/_jcr_content/renditions/original./malignant-melanoma-may-2020.pdf" target="_blank">Source</a>

```{r}
plot_km_age <- function(cutoff_age_arr) {
  for (a in cutoff_age_arr) {
    lt_age = paste("Less than ", a, sep="")
    ov_age = paste("Over ", a, sep="")
    mm_age_adj = melanoma.df %>% mutate(AG = ifelse((age < a), lt_age, ov_age),
                 AG = factor(AG))
    
    km_fit_age <- survfit(
      Surv(time, status_general)~AG, 
      data=mm_age_adj
    )
    
    plot_title = "Kaplan Meier plot for the survival of Maglinant Melanoma(MM) by age"
    km_age_plot <- autoplot(km_fit_age)
    km_age_plot <- km_age_plot + 
      ggtitle(plot_title) +
      labs(x = xlabel, y = ylabel) +
      guides(fill=FALSE) +
      labs(colour = "AGE") +
      scale_color_manual(labels = c(lt_age, ov_age), values = c(1, 2))
    print(km_age_plot) 
  }
}

cutoff_ages <- c(40,50,60,70)
plot_km_age(cutoff_ages)
```

## Model fit and variable selection  - Cox proportional-hazards model analysis

Testing proportional hazard assumption

```{r}
par(mfrow=c(2,3))
death_status <-  melanoma.df$status_general
fit.cox = coxph(Surv(time,death_status)~factor(sex)+age+thickness+year+factor(ulcer),method=c("breslow"), na.action=na.exclude, data=melanoma.df)
fit.cox.zph = cox.zph(fit.cox)
summary(fit.cox.zph)
plot(fit.cox.zph)

fit.cox.zph
```

The component `ulcer` with chi-square value of 4.320, under df = 1, has a p-value of 0.038. This implies that Proportional Hazards assumption does not hold. So we need to stratify our ulceration component, and check it again.

```{r}
par(mfrow=c(2,3))
fit.cox = coxph(Surv(time,death_status)~factor(sex)+age+thickness+year+strata(factor(ulcer)),method=c("breslow"), na.action=na.exclude, data=melanoma.df)
fit.cox.zph = cox.zph(fit.cox)
summary(fit.cox.zph)
plot(fit.cox.zph)
abline(h=0, lty=3)
fit.cox.zph
```

We perform local tests for each explanatory factor in the model by applying Wald test.

```{r, echo=FALSE}
coxph.localtest<-function(coxph.fit, df){
  coef<-coxph.fit$coef
  var<-coxph.fit$var
  loglik<-coxph.fit$loglik[2]
  p<-length(coef)
  AIC<- -2*loglik+2*p    #Using the formula on p.277;
  var.mat<-solve(as.matrix(var[(p-df+1):p, (p-df+1):p]))
  coe.mat<-matrix(c(coef[(p-df+1):p]), df, 1)
  WaldChiSq<-t(coe.mat)%*%var.mat%*%coe.mat
  pvalue<-1-pchisq(WaldChiSq,df)
  results<-c(df, WaldChiSq, pvalue, AIC)
  list(results)
}

options(width=60,length=200,digits=5) 

Table_checkup<-matrix(0,3,4)

coxph.sex<-coxph(Surv(time,death_status)~strata(factor(ulcer))+thickness+factor(sex), method=c("breslow"), data=melanoma.df)
coxph.age<-coxph(Surv(time,death_status)~strata(factor(ulcer))+thickness+age, method=c("breslow"), data=melanoma.df)
coxph.year <- coxph(Surv(time,death_status)~strata(factor(ulcer))+thickness+year, method=c("breslow"), data=melanoma.df)

Table_checkup[1,]<-c(coxph.localtest(coxph.sex, df=1)[[1]])
Table_checkup[2,]<-c(coxph.localtest(coxph.age, df=1)[[1]])
Table_checkup[3,]<-c(coxph.localtest(coxph.year, df=1)[[1]])

cat("Table 8.6: Local test for possible confounders, adjusted for factors of gender, age, thickness and ulceration", "\n")
cat("    DF, Wald Chi-Squre, p-value, AIC:", "\n")
print(Table_checkup)
```

The p_values of sex, and year are more than 0.05. We fail to reject Ho and said sex, and year is not effective. So our finalized model is the one below.

```{r, echo=FALSE}
coxph_final<-coxph(Surv(time,death_status)~strata(factor(ulcer))+thickness+age, method=c("breslow"), data=melanoma.df)

cat("Table 8.9: Analysis of Variance Table for the Final Model for the survival of Malignant Melanoma", "\n")
print(coxph_final)
```

We double check our decision by taking the forward and backward AIC method. Our results turn out same which suggests rejection to year, and sex.

#### Backward AIC method

```{r}
fit.cox <- coxph(Surv(time,death_status)~1,data=melanoma.df)
stepAIC(fit.cox,  ditrection="backward",trace=-1) 
```

#### Forward AIC method

```{r}
stepAIC(fit.cox, direction="forward", 
        scope=list(lower=fit.cox, upper=~factor(sex)+age+thickness+year+factor(ulcer)))
```

So our model without adjusting for competing risks is <br/>
>formula = Surv(time, death_status) ~ factor(ulcer) + thickness + age

## Cox proportional-hazards model adjusted for competing risks

#### Cox proportional-hazards model classification 2
We defined this class to be such that the failure time is the MM mortality with 1 as death by MM, 2/3 are alive regardless of other causes of death.

```{r, echo=FALSE}
explanatory = c("age", "sex", "thickness", "ulcer")
mm_class = "Surv(time, status_mm)"
cen_class = "Surv(time, status_cen)"
```

```{r}
melanoma.df %>%coxphmulti(mm_class, explanatory)
```

#### Cox proportional-hazards model classification 3 - censoring applied

We defined this class to be such that the failure time with censoring adjust for competing risks where 2 represents alive, 1 is death by MM adjusted for 3 which is death by other causes.

```{r}
melanoma.df%>%crrmulti(cen_class, explanatory)
```

# Reference

* <a href="https://acsjournals.onlinelibrary.wiley.com/doi/pdf/10.1002/1097-0142(19820601)49:11%3C2414::AID-CNCR2820491132%3E3.0.CO;2-V" target="_blank">  Drzewieki K. T., Andersen P., Survival with Malignant Melanoma: A Regression Analysis of Prognostic Factors . American Cancer Society 1988 </a>

* <a href="https://www.munichre.com/content/dam/munichre/marc/pdf/malignant-melanoma-may-2020.pdf/_jcr_content/renditions/original./malignant-melanoma-may-2020.pdf" target="_blank">  Drzewieki K. T., Andersen P., Malignant
Melanoma .  Munich American Reassurance Company 2020 </a>